Runtime-Assisted Convergence in Replicated Data Types

### 2022-03-15 motivation.tex
Incorporating reviewer comments

\noindent\paragraph{Key observations guiding the solution}
Fortunately, there is a way out of this problem and the solution
requires a switch in the perspective of replication from an
operation-centric view characterized by explicit effects to a
state-centric view characterized by explicit state merges. The
motivation for this switch is the key observation that it is possible
to induce an ordering over \emph{states} of replicas even as their
operations and effects remain unordered, as long as such states are
\emph{mergeable}. The ordering is enforced by synchronizing merges in
the background while the replica-local execution progresses
unhindered. Consequently, eventual convergence can be guaranteed
\emph{without} impacting the user-perceived latency of the operations.
Moreover, working with states instead of operations frees the latter
from having to conform to algebraic laws such as commutativity.  Put
together, this means that the clients of a replicated data type can
perform any action allowed by the sequential version of the type and
immediately see the results of their actions on the local replica
while being safe in the knowledge that replica states will eventually
converge.

@@

\noindent\paragraph{A Mergeable Replicated Set} To demonstrate the
aforementioned intuitions we reconsider the executions from
Fig.~\ref{fig:crdt-execs}, this time in the state-centric replication
model. The cornerstone of the state-centric model is a three-way
\C{merge} function that merges concurrent versions of the state in
presence of their (lowest) common ancestor version. In our running
example, the state is a value of type \C{Set.t}, hence \C{Set.merge}
function would have the type signature:
\begin{center}
\C{Set.merge} : \C{Set.t} $\rightarrow$ \C{Set.t} $\rightarrow$ \C{Set.t}
$\rightarrow$ \C{Set.t}
\end{center}
The three arguments of \C{merge} correspond to the lowest common
ancestor (LCA) version and the two concurrent versions that
independently evolved from the LCA version. The LCA is a causal
ancestor of concurrent versions, hence causal consistency is built
into the replication model. The result of set merge, intuitively, must
contain the common elements in the two concurrent versions along with
any newly added elements in either versions. Concretely:
\begin{ocaml}
  let merge s(*lca*) s1 s2 = 
              (s1 $\cap$ s2) $\cup$ (s1 - s) $\cup$ (s2 - s)
\end{ocaml}
Equipped with the above set merge function, we can now consider the
equivalent execution of Fig.~\ref{fig:crdt-exec-1} in state-centric
model. The execution is shown in Fig.~\ref{fig:mrdt-exec-1}. The
initial version on all three replicas ($v_0$) is the singleton set
$\{e\}$. Applying operations to replicas creates new versions, e.g.,
$v_1$ on $R_3$, and $v_2$ and $v_3$ on $R_2$. Changes can be propagated by merging versions, e.g.,
version $v_2$ on $R_2$ is a result of merging $v_1$ into $v_0$ for
which $v_0$ serves as the lowest common ancestor (LCA)
  \footnote{Versions $v_0$ and $v_1$ are not concurrent as the
  former is an ancestor of the latter. Merging $v_1$ into $v_0$ is
  nonetheless possible as $v_1$ is ahead of $v_0$ in causal order. In
  Git parlance this is a \emph{fast forward} merge.}.
Likewise $v_4$ on $R_1$ is created by merging $v_2$ into $v_0$ in
presence of their LCA $v_0$. The next version $v_5$ on $R_1$ is the
result of merging $R_2$'s $v_3$ into $R_1$'s $v_4$. The LCA for this
merge is $v_2$. By the end of the execution, versions $v_5$ and $v_3$
on $R_1$ and $R_2$ (resp.) have witnessed the same set of operations,
hence are in agreement.

@@

Note that the anamolous execution in Fig.~\ref{fig:mrdt-exec-2} could
have been avoided had the merges between $R_1$ and $R_2$ been
linearized. Fig.~\ref{fig:mrdt-exec-3} shows an execution that only
slightly differs from the one in Fig.~\ref{fig:mrdt-exec-2}. The
difference is that the merges in Fig.~\ref{fig:mrdt-exec-3} happen
linearly: first $R_1$ is merged into $R_2$ (bringing $R_1$'s
\C{remove}), then $R_2$ into $R_1$ (bringing $R_2$'s \C{remove}),
followed again by $R_2$ into $R_1$ (bringing $R_2$'s \C{add}). As a
result of such linearization, the final versions on $R_1$ and $R_2$
converge to the singleton set $\{e\}$. Note that there exist other
linearizations of merges; for instance, $R_1 \,\rightarrow\, R_2$
merge could be ordered between the two $R_2 \rightarrow R_1$ merges.
However, all linearizations result in the same final state $\{e\}$.
Another important point to note is that only the merges are
linearized; not the entire execution. In Fig.~\ref{fig:mrdt-exec-3},
both $R_1$ and $R_2$'s \C{remove}s remove the same element $e$, which
is not possible in a linearized execution. Leaving the execution
unconstrained is crucial to satisfy the limits imposed by the CAP
theorem on a highly-available partition-tolerant distributed system.

In the context of Fig.~\ref{fig:mrdt-exec-3}, it is quite clear what
linearization of merges means and how to enforce it via
synchronization (for e.g., wrapping each merge within a global lock).
In general however, the semantics of merge linearization isn't as cut
and dried. For instance, consider the execution in
Fig.~\ref{fig:mrdt-exec-4}. The four replicas involved in the
execution start with version $v_0 \,=\, \{e\}$. The replicas perform
local operations as shown in the figure to make versions $v_1$ to
$v_3$. Next they perform a series of merges to propagate local
changes. The merges can be ordered in time as following: first $R_2
\rightarrow R_1$ (merging $v_1$), then $R_3 \rightarrow R_1$ twice
(merging $v_2$ and $v_3$ resp.), then $R_3 \rightarrow R_4$ (merging
$v_3)$, and finally $R_2 \rightarrow R_4$ (merging $v_1$). These
merges collectively propagate the effects of \C{add} and \C{remove}
operations to $R_1$ and $R_2$. And despite being ordered in time, the
they nonetheless result in a divergent execution ($v_5 \neq v_7$). The
problem here is that, although merges are executed linearly, the
execution graph does not reflect this linearity; merges of that end in
$R_1$ in Fig.~\ref{fig:mrdt-exec-4} are effectively concurrent with
those that end in $R_4$. This shows that simply synchronizing the
execution of merges does not necessarily result in convergence.

Our key insight to overcome this impasse is a \emph{well-formedness}
condition on execution graphs that ensures convergence of final
states. To understand well-formedness, let us contrast the bad
executions in Figs.~\ref{fig:mrdt-exec-2} and~\ref{fig:mrdt-exec-4}
against the good execution in Fig.~\ref{fig:mrdt-exec-3}. Observe that
in Fig.~\ref{fig:mrdt-exec-2}, \emph{every} pair of concurrent
versions on $R_1$ and $R_2$ have a unique lowest common ancestor
(LCA). For instance, LCA of $(v_5,v_6)$ is $v_6$, $(v_5,v_4)$ is
$v_3$, $(v_1,v_2)$ is $v_0$ and so on. By contrast in
Fig.~\ref{fig:mrdt-exec-2}, versions $v_5$ and $v_6$ have two LCAs,
namely $v_2$ and $v_3$. Both these versions are common ancestors of
$v_5$ and $v_6$, and both are \emph{lowest} in the sense that there do
not exist versions lower (in the execution graph) than $v_2$ and $v_3$
that are also common ancestors of $v_5$ and $v_6$. Likewise in
Fig.~\ref{fig:mrdt-exec-4}, versions $v_7$ and $v_9$ have two LCAs --
$v_2$ and $v_4$. Multiple LCAs is an indication that there exist
merges prior in the execution graph that are effectively concurrent.
Considering that our approach to convergence in state-centric model
crucially relies on linearity of merges, presence of multiple LCAs
opens up the possibility of divergence. We therefore define
well-formed execution graphs as those where LCAs for every pair of
concurrent versions is unique. As we prove in
Sec.~\ref{sec:abstract-sem}, enforcing this structural
well-formedness condition indeed guarantees the convergence of
distributed executions. In Sec.~\ref{sec:concrete-sem}, we formalize a
distributed runtime called \quark whose primary purpose is to enforce
the well-formedness condition on distributed executions. In
Sec.~\ref{sec:implementation}, we describe its implementation. The
runtime makes it possible to automatically enforce convergence,
letting us promote ordinary sequential data types to convergent
replicated types by simply equipping them with a merge function.
Thus, extending the \C{Set} interface of Fig.~\ref{fig:ocaml-set} with
the two-line \C{Set.merge} function shown above is sufficient to
obtain a convergent replicated set data type.

>>>>

@@

\noindent\paragraph{The MRDT Approach} Mergeable Replicated Data Types
(MRDTs) implement a state-centric model of replication that is
inspired by the Git version control system. Like in Git, the state
evolves in terms of versions, and concurrent versions of the state can
be merged. The semantics of merge depends on the type of the state, so
each MRDT is required to be equipped with a three-way \C{merge}
function that merges concurrent versions of that type in presence of
their (lowest) common ancestor version. In our running example, the
state is a value of type \C{Set.t}, hence \C{Set.merge} function would
have the type signature:
\begin{center}
\C{Set.merge} : \C{Set.t} $\rightarrow$ \C{Set.t} $\rightarrow$ \C{Set.t}
$\rightarrow$ \C{Set.t}
\end{center}
The three arguments of \C{merge} correspond to the lowest common
ancestor (LCA) version and the two concurrent versions that
independently evolved from the LCA version. The LCA is a causal
ancestor of concurrent versions, hence causal consistency is built
into the replication model. The result of set merge, intuitively, must
contain the common elements in the two concurrent versions along with
any newly added elements in either versions. Concretely:
\begin{ocaml}
  let merge s(*lca*) s1 s2 = 
              (s1 $\cap$ s2) $\cup$ (s1 - s) $\cup$ (s2 - s)
\end{ocaml}
Extending \C{Set} ADT with the above merge function results in a
\C{Set} MRDT.

Let us now reconsider the executions in Fig.~\ref{fig:crdt-execs},
this time on \C{Set} MRDT. The equivalent executions are shown in
Fig.~\ref{fig:mrdt-execs}. Due to causal consistency being built into
the model, the divergent execution of Fig.~\ref{fig:crdt-exec-1} is
preempted in favor of the convergent execution shown in
Fig.~\ref{fig:mrdt-exec-1}. The execution manifests as follows.  The
initial version ($v_0$) on all three replicas  is the singleton set
$\{e\}$. Applying operations to replicas creates new versions, e.g.,
$v_1$ on $R_3$, and $v_2$ and $v_3$ on $R_2$. Changes can be
propagated by merging versions, e.g., version $v_2$ on $R_2$ is a
result of merging $v_1$ into $v_0$ for which $v_0$ serves as the
lowest common ancestor (LCA)
  \footnote{Versions $v_0$ and $v_1$ are not concurrent as the
  former is an ancestor of the latter. Merging $v_1$ into $v_0$ is
  nonetheless possible as $v_1$ is ahead of $v_0$ in causal order. In
  Git parlance this is a \emph{fast forward} merge.}.
Likewise $v_4$ on $R_1$ is created by merging $v_2$ into $v_0$ in
presence of their LCA $v_0$. The next version $v_5$ on $R_1$ is the
result of merging $R_2$'s $v_3$ into $R_1$'s $v_4$. The LCA for this
merge is $v_2$. By the end of the execution, versions $v_5$ and $v_3$
on $R_1$ and $R_2$ (resp.) have witnessed the same set of operations,
hence are in agreement.

@@

Note that the anamolous execution in Fig.~\ref{fig:mrdt-exec-2} could
have been avoided had the merges between $R_1$ and $R_2$ been
linearized. Fig.~\ref{fig:mrdt-exec-3} shows an execution that only
slightly differs from the one in Fig.~\ref{fig:mrdt-exec-2}. The
difference is that the merges in Fig.~\ref{fig:mrdt-exec-3} happen
linearly: first $R_1$ is merged into $R_2$ (bringing $R_1$'s
\C{remove}), then $R_2$ into $R_1$ (bringing $R_2$'s \C{remove}),
followed again by $R_2$ into $R_1$ (bringing $R_2$'s \C{add}). As a
result of such linearization, the final versions on $R_1$ and $R_2$
converge to the singleton set $\{e\}$. Note that there exist other
linearizations of merges; for instance, $R_1 \,\rightarrow\, R_2$
merge could be ordered between the two $R_2 \rightarrow R_1$ merges.
However, all linearizations result in the same final state $\{e\}$.
Another important point to note is that only the merges are
linearized; not the entire execution. In Fig.~\ref{fig:mrdt-exec-3},
both $R_1$ and $R_2$'s \C{remove}s remove the same element $e$, which
is not possible in a linearized execution. %Leaving the execution
%unconstrained is crucial to satisfy the limits imposed by the CAP
%theorem on a highly-available partition-tolerant distributed system.

In the context of Fig.~\ref{fig:mrdt-exec-3}, it is quite clear what
linearization of merges means and how to enforce it via
synchronization (for e.g., wrapping each merge within a global lock).
In general however, the semantics of merge linearization isn't as cut
and dried. For instance, consider the execution in
Fig.~\ref{fig:mrdt-exec-4}. The four replicas involved in the
execution start with version $v_0 \,=\, \{e\}$. The replicas perform
local operations as shown in the figure to make versions $v_1$ to
$v_3$. Next they perform a series of merges to propagate local
changes. The merges can be ordered in time as following: first $R_2
\rightarrow R_1$ (merging $v_1$), then $R_3 \rightarrow R_1$ twice
(merging $v_2$ and $v_3$ resp.), then $R_3 \rightarrow R_4$ (merging
$v_3)$, and finally $R_2 \rightarrow R_4$ (merging $v_1$). These
merges collectively propagate the effects of \C{add} and \C{remove}
operations to $R_1$ and $R_2$. And despite being ordered in time,
they nonetheless result in a divergent execution ($v_5 \neq v_7$). The
problem here is that, although merges are executed linearly, the
execution graph does not reflect this linearity; merges that end in
$R_1$ in Fig.~\ref{fig:mrdt-exec-4} are effectively concurrent with
those that end in $R_4$. This shows that simply synchronizing the
execution of merges does \emph{not} result in convergence.

Our key insight to overcome this impasse is a novel \emph{well-formedness}
condition on execution graphs that ensures convergence of final
states. To understand well-formedness, let us contrast the bad
executions in Figs.~\ref{fig:mrdt-exec-2} and~\ref{fig:mrdt-exec-4}
against the good execution in Fig.~\ref{fig:mrdt-exec-3}. Observe that
in Fig.~\ref{fig:mrdt-exec-3}, \emph{every} pair of concurrent
versions on $R_1$ and $R_2$ have a unique lowest common ancestor
(LCA). For instance, LCA of $(v_5,v_6)$ is $v_5$, $(v_5,v_4)$ is
$v_3$, $(v_1,v_2)$ is $v_0$ and so on. By contrast in
Fig.~\ref{fig:mrdt-exec-2}, versions $v_5$ and $v_6$ have two LCAs,
namely $v_2$ and $v_3$. Both these versions are common ancestors of
$v_5$ and $v_6$, and both are \emph{lowest} in the sense that there do
not exist versions lower (in the execution graph) than $v_2$ and $v_3$
that are also common ancestors of $v_5$ and $v_6$. Likewise in
Fig.~\ref{fig:mrdt-exec-4}, versions $v_6$ and $v_8$ have two LCAs --
$v_1$ and $v_3$. Multiple LCAs is an indication that there exist
merges prior in the execution graph that are effectively concurrent.
Considering that our approach to convergence in state-centric model
crucially relies on linearity of merges, presence of multiple LCAs
opens up the possibility of divergence. We therefore define
well-formed execution graphs as those where LCAs for every pair of
concurrent versions is unique. As we prove in
Sec.~\ref{sec:abstract-sem}, enforcing this structural
well-formedness condition indeed guarantees the convergence of
distributed executions. In Sec.~\ref{sec:concrete-sem}, we formalize a
distributed runtime called \quark whose primary purpose is to enforce
the well-formedness condition on distributed executions. In
Sec.~\ref{sec:implementation}, we describe its implementation. The
runtime makes it possible to automatically enforce convergence,
letting us promote ordinary sequential data types to convergent
replicated types by simply equipping them with a merge function.
Thus, extending the \C{Set} interface of Fig.~\ref{fig:ocaml-set} with
the two-line \C{Set.merge} function shown above is sufficient to
obtain a convergent replicated set data type.

### 2022-03-15 quark.tex
Edits
Incorporating reviewer comments

%% For double-blind review submission, w/o CCS and ACM Reference (max submission space)
\documentclass[sigplan,10pt,review,anonymous]{acmart}
\settopmatter{printfolios=true,printccs=false,printacmref=false}
%% For final camera-ready submission, w/ required CCS and ACM Reference
%\documentclass[acmsmall]{acmart}\settopmatter{}

@@

%% Journal information
%% Supplied to authors by publisher for camera-ready submission;
%% use defaults for review submission.
\acmJournal{PACMPL}
\acmVolume{1}
\acmNumber{POPL} % CONF = POPL or ICFP or OOPSLA
\acmArticle{1}
\acmYear{2022}
\acmMonth{1}
\acmDOI{} % \acmDOI{10.1145/nnnnnnn.nnnnnnn}
\startPage{1}

@@

%% Author with single affiliation.
\author{First1 Last1}
\authornote{with author1 note}          %% \authornote is optional;
                                        %% can be repeated if necessary
\orcid{nnnn-nnnn-nnnn-nnnn}             %% \orcid is optional
\affiliation{
  \position{Position1}
  \department{Department1}              %% \department is recommended
  \institution{Institution1}            %% \institution is required
  \streetaddress{Street1 Address1}
  \city{City1}
  \state{State1}
  \postcode{Post-Code1}
  \country{Country1}                    %% \country is recommended
}
\email{first1.last1@inst1.edu}          %% \email is recommended

%% Author with two affiliations and emails.
\author{First2 Last2}
\authornote{with author2 note}          %% \authornote is optional;
                                        %% can be repeated if necessary
\orcid{nnnn-nnnn-nnnn-nnnn}             %% \orcid is optional
\affiliation{
  \position{Position2a}
  \department{Department2a}             %% \department is recommended
  \institution{Institution2a}           %% \institution is required
  \streetaddress{Street2a Address2a}
  \city{City2a}
  \state{State2a}
  \postcode{Post-Code2a}
  \country{Country2a}                   %% \country is recommended
}
\email{first2.last2@inst2a.com}         %% \email is recommended
\affiliation{
  \position{Position2b}
  \department{Department2b}             %% \department is recommended
  \institution{Institution2b}           %% \institution is required
  \streetaddress{Street3b Address2b}
  \city{City2b}
  \state{State2b}
  \postcode{Post-Code2b}
  \country{Country2b}                   %% \country is recommended
}
\email{first2.last2@inst2b.org}         %% \email is recommended

%% Abstract
%% Note: \begin{abstract}...\end{abstract} environment must come
%% before \maketitle command
\begin{abstract}

  We propose a runtime-assisted approach to enforce convergence in
  distributed executions of replicated data types. The key
  distinguishing aspect of our approach is that it guarantees
  convergence \emph{unconditionally} -- without requiring data
  type operations to satisfy algebraic laws such as commutativity and
  idempotence. Consequently, programmers are no longer obligated to
  prove convergence on a per-type basis. Moreover, our approach lets
  sequential data types be reused in a distributed setting by
  extending their implementations rather than refactoring them.
  The novel component of our approach is a distributed runtime that
  orchestrates \emph{well-formed } executions that are guaranteed to
  converge. Despite the utilization of a runtime, our approach comes
  at no additional cost of latency and availability. We implement our
  approach in a system called \quark and evaluate its performance over
  a large-scale collaborative editing application. 
\end{abstract}

%% 2012 ACM Computing Classification System (CSS) concepts
%% Generate at 'http://dl.acm.org/ccs/ccs.cfm'.
\begin{CCSXML}
<ccs2012>
<concept>
<concept_id>10011007.10011006.10011008</concept_id>
<concept_desc>Software and its engineering~General programming languages</concept_desc>
<concept_significance>500</concept_significance>
</concept>
<concept>
<concept_id>10003456.10003457.10003521.10003525</concept_id>
<concept_desc>Social and professional topics~History of programming languages</concept_desc>
<concept_significance>300</concept_significance>
</concept>
</ccs2012>
\end{CCSXML}

\ccsdesc[500]{Software and its engineering~General programming languages}
\ccsdesc[300]{Social and professional topics~History of programming languages}
%% End of generated code

%% Keywords
%% comma separated list
\keywords{keyword1, keyword2, keyword3}  %% \keywords are mandatory in final camera-ready submission

@@

%% Acknowledgments
\begin{acks}                            %% acks environment is optional
                                        %% contents suppressed with 'anonymous'
  %% Commands \grantsponsor{<sponsorID>}{<name>}{<url>} and
  %% \grantnum[<url>]{<sponsorID>}{<number>} should be used to
  %% acknowledge financial support and will be used by metadata
  %% extraction tools.
  This material is based upon work supported by the
  \grantsponsor{GS100000001}{National Science
    Foundation}{http://dx.doi.org/10.13039/100000001} under Grant
  No.~\grantnum{GS100000001}{nnnnnnn} and Grant
  No.~\grantnum{GS100000001}{mmmmmmm}.  Any opinions, findings, and
  conclusions or recommendations expressed in this material are those
  of the author and do not necessarily reflect the views of the
  National Science Foundation.
\end{acks}

@@

Text of appendix \ldots

>>>>

%% For double-blind review submission, w/o CCS and ACM Reference (max submission space)
\documentclass[sigplan,screen]{acmart}
%\settopmatter{printfolios=true,printccs=false,printacmref=false}
%% For final camera-ready submission, w/ required CCS and ACM Reference
%\documentclass[acmsmall]{acmart}\settopmatter{}

@@

%% Journal information
%% Supplied to authors by publisher for camera-ready submission;
%% use defaults for review submission.
%%\acmJournal{PACMPL}
%%\acmVolume{1}
%%\acmNumber{POPL} % CONF = POPL or ICFP or OOPSLA
%%\acmArticle{1}
%%\acmYear{2022}
%%\acmMonth{1}
%%\acmDOI{} % \acmDOI{10.1145/nnnnnnn.nnnnnnn}
%%\startPage{1}
\acmConference[PLDI]{ACM Conference}{June 2022}{San Diego, USA}%

@@

%% Author with single affiliation.
\author{Gowtham Kaki}
%\authornote{with author1 note}          %% \authornote is optional;
                                        %% can be repeated if necessary
%\orcid{nnnn-nnnn-nnnn-nnnn}             %% \orcid is optional
\affiliation{
%  \position{Position1}
%  \department{Department1}              %% \department is recommended
  \institution{University of Colorado Boulder}            %% \institution is required
%  \streetaddress{Street1 Address1}
%  \city{City1}
%  \state{State1}
%  \postcode{Post-Code1}
  \country{USA}                    %% \country is recommended
}
%\email{first1.last1@inst1.edu}          %% \email is recommended

%% Second Author
\author{Prasanth Prahladan}
%\authornote{with author2 note}          %% \authornote is optional;
                                        %% can be repeated if necessary
%\orcid{nnnn-nnnn-nnnn-nnnn}             %% \orcid is optional
\affiliation{
%  \position{Position2a}
%  \department{Department2a}             %% \department is recommended
  \institution{University of Colorado Boulder}           %% \institution is required
%  \streetaddress{Street2a Address2a}
%  \city{City2a}
%  \state{State2a}
%  \postcode{Post-Code2a}
  \country{USA}                   %% \country is recommended
}
%\email{first2.last2@inst2a.com}         %% \email is recommended

%% Third Author
\author{Nicholas Lewchenko}
%\authornote{with author2 note}          %% \authornote is optional;
                                        %% can be repeated if necessary
%\orcid{nnnn-nnnn-nnnn-nnnn}             %% \orcid is optional
\affiliation{
%  \position{Position2a}
%  \department{Department2a}             %% \department is recommended
  \institution{University of Colorado Boulder}           %% \institution is required
%  \streetaddress{Street2a Address2a}
%  \city{City2a}
%  \state{State2a}
%  \postcode{Post-Code2a}
  \country{USA}                   %% \country is recommended
}
%

%% Abstract
%% Note: \begin{abstract}...\end{abstract} environment must come
%% before \maketitle command
\begin{abstract}
  We propose a runtime-assisted approach to enforce convergence in
  distributed executions of replicated data types. The key
  distinguishing aspect of our approach is that it guarantees
  convergence \emph{unconditionally} -- without requiring data
  type operations to satisfy algebraic laws such as commutativity and
  idempotence. Consequently, programmers are no longer obligated to
  prove convergence on a per-type basis. Moreover, our approach lets
  sequential data types be reused in a distributed setting by
  extending their implementations rather than refactoring them.
  The novel component of our approach is a distributed runtime that
  orchestrates \emph{well-formed } executions that are guaranteed to
  converge. Despite the utilization of a runtime, our approach comes
  at no additional cost of latency and availability. Instead, we
  introduce a novel tradeoff against a metric called \emph{staleness},
  which roughly corresponds to the time taken for replicas to
  converge. We implement our approach in a system called \quark and
  conduct a thorough evaluation of its tradeoffs in the context of two
  case studies.
\end{abstract}

%% 2012 ACM Computing Classification System (CSS) concepts
%% Generate at 'http://dl.acm.org/ccs/ccs.cfm'.
\begin{CCSXML}
<ccs2012>
<concept>
<concept_id>10010147.10010919.10010177</concept_id>
<concept_desc>Computing methodologies~Distributed programming languages</concept_desc>
<concept_significance>500</concept_significance>
</concept>
<concept>
<concept_id>10010520.10010575.10010578</concept_id>
<concept_desc>Computer systems organization~Availability</concept_desc>
<concept_significance>500</concept_significance>
</concept>
<concept>
<concept_id>10011007.10011074.10011099.10011692</concept_id>
<concept_desc>Software and its engineering~Formal software verification</concept_desc>
<concept_significance>300</concept_significance>
</concept>
</ccs2012>
\end{CCSXML}

\ccsdesc[500]{Computing methodologies~Distributed programming languages}
\ccsdesc[500]{Computer systems organization~Availability}
\ccsdesc[300]{Software and its engineering~Formal software verification}
%% End of generated code

%% Keywords
%% comma separated list
\keywords{Replication, MRDT, CRDT, Runtime, Convergence, Collaborative
Editing}  %% \keywords are mandatory in final camera-ready submission

@@

%% Acknowledgments
%\begin{acks}                            %% acks environment is optional
                                        %% contents suppressed with 'anonymous'
  %% Commands \grantsponsor{<sponsorID>}{<name>}{<url>} and
  %% \grantnum[<url>]{<sponsorID>}{<number>} should be used to
  %% acknowledge financial support and will be used by metadata
  %% extraction tools.
%\end{acks}

@@
